---
title: "R Notebook"
output: html_notebook
---

# Setup

## Load libraries

```{r}
quiet <- suppressPackageStartupMessages

quiet(library(tidyverse))
quiet(library(openxlsx))
quiet(library(DESeq2))
quiet(library(janitor))

theme_set(theme_classic())
```

## Util functions

```{r}
pick <- function(rdf, target_col, as_char=FALSE) {
  target_col <- enquo(target_col)
  raw_col <- rdf %>% dplyr::select(!!target_col) %>% unlist() %>% unname()
  if (as_char) {
    raw_col %>% as.character()
  }
  else {
    raw_col
  }
}
```

## Load data

### Parse from raw data

#### Setup sampling

```{r}
design_seq_df <- openxlsx::read.xlsx("data/GTEx_pancreas_liver_images_liverfat_pancreasfat_seq.xlsx") %>% janitor::clean_names()
design_pat_df <- openxlsx::read.xlsx("data/GTEx_pancreas_liver_images_liverfat_pancreasfat.xlsx")

validation_samples <- sample(nrow(design_seq_df), replace=FALSE, size = 20)

design_seq_df$is_validation_sample <- seq_len(nrow(design_seq_df)) %in% validation_samples

design_seq_df

write.xlsx(design_seq_df, file = "data/design_seq_w_valid.xlsx")

```

#### Setup design

```{r}
design_seq_df <- openxlsx::read.xlsx("data/design_seq_w_valid.xlsx")

cor(design_seq_df$fat_percentage_liver, design_seq_df$fat_percentage_pancreas)
cor.test(design_seq_df$fat_percentage_liver, design_seq_df$fat_percentage_pancreas)

pancreas_cols <- colnames(design_seq_df[grepl("_pancreas$", colnames(design_seq_df))])
liver_cols <- colnames(design_seq_df[grepl("_liver$", colnames(design_seq_df))])
tissue_cols_names <- pancreas_cols %>% gsub("_pancreas", "", .)
other_cols <- colnames(design_seq_df)[!colnames(design_seq_df) %in% c(pancreas_cols, liver_cols)]
non_tissue_design_data <- design_seq_df[, !(colnames(design_seq_df) %in% c(pancreas_cols, liver_cols))] %>%
  mutate(id=paste("patient", seq_len(n()), sep="_")) %>% 
  dplyr::select(id, everything())

# pancreas_fat_cut <- 10
pancreas_fat_cut <- median(design_seq_df$fat_percentage_pancreas)

pancreas_design_distinct <- cbind(
  non_tissue_design_data, 
  design_seq_df[, pancreas_cols] %>% 
    `colnames<-`(tissue_cols_names) %>% mutate(source="pancreas")) %>% 
  distinct(sampid, .keep_all=TRUE) %>%
  mutate(over_fat_cutoff = fat_percentage > median(fat_percentage))

# liver_fat_cut <- 35
liver_fat_cut <- median(design_seq_df$fat_percentage_liver)
liver_design_distinct <- cbind(
  non_tissue_design_data, design_seq_df[, liver_cols] %>% 
    `colnames<-`(tissue_cols_names) %>% 
    mutate(source="liver")) %>%
  distinct(sampid, .keep_all=TRUE) %>%
  mutate(over_fat_cutoff = fat_percentage > median(fat_percentage))

ddf <- rbind(pancreas_design_distinct, liver_design_distinct)

message("Any duplicated samples present: ", all(duplicated(ddf$sampid)))
```

#### Parse further sample information

```{r}
sample_attributes <- read_tsv("data/GTEx_Analysis_v8_Annotations_SampleAttributesDS.txt") %>% janitor::clean_names()
extended_ddf <- ddf %>% inner_join(sample_attributes, by="sampid")
write_tsv(extended_ddf, path="data/count_matrix_target_subset.design.tsv")


```

#### Include the sample information matrix

```{r eval=FALSE}
# raw_df <- read_tsv("data/GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_reads.gct", skip = 2)
# 
# length(unique(ddf$SAMPID))
# length(ddf$SAMPID)
# 
# rdf <- raw_df %>% dplyr::select(Name, Description, any_of(ddf$sampid))
# write_tsv(rdf, path = "data/count_matrix_target_subset.tsv.gz")
```



### Parse from preparsed data

```{r}
rdf <- read_tsv("data/count_matrix_target_subset.tsv.gz")
ddf_all <- read_tsv("data/count_matrix_target_subset.design.tsv") %>% arrange(sampid)

ddf_all

ddf <- ddf_all %>% dplyr::filter(sampid %in% colnames(rdf))

plt <- ggpubr::ggarrange(
  ddf %>% 
    filter(source == "liver") %>%
    ggplot(aes(x=fat_percentage, fill=fat_percentage<liver_fat_cut)) + geom_histogram() + xlim(0, 100) + ggtitle("Liver fat percentage"),
  ddf %>%
    filter(source == "pancreas") %>%
    ggplot(aes(x=fat_percentage, fill=fat_percentage<pancreas_fat_cut)) + geom_histogram() + xlim(0, 100) + ggtitle("Pancreas fat percentage"),
  nrow=2, ncol=1
)
plt
ggsave(plt, filename = "output/fat_percentages_high.png")
```

### Overview the input data

Note that 25 out of 204 samples does not seem to be present in the count matrix.

Four are missing in both types of tissues.

There seem to be duplicate IDs, but that could be due to multiple tissue sample taken from the same person of one type of tissue.

So this could maybe safely be omitted?

But need to think carefully about the statistics!

```{r}
message("Of all samples, how many are present in count data")
table((ddf_all %>% pick(sampid)) %in% colnames(rdf))
message("Of liver samples, how many are present in count data")
present_in_liver <- (ddf_all %>% filter(source == "pancreas") %>% pick(sampid)) %in% colnames(rdf)
table(present_in_liver)
message("Of pancreas samples, how many are present in count data")
present_in_pancreas <- (ddf_all %>% filter(source == "liver") %>% pick(sampid)) %in% colnames(rdf)
table(present_in_pancreas)
message("Of these, how many are missing simultaneously in both?")
table(present_in_liver | present_in_pancreas)
```

# DESeq2 analysis

## Utility functions

```{r}
make_deseq_data <- function(rdf, sample_cols) {
  sdf_dataframe <- rdf %>% dplyr::select(all_of(sort(sample_cols))) %>% as.data.frame()
  rownames(sdf_dataframe) <- rdf$Name
  sdf_dataframe
}

make_deseq_design <- function (ddf_raw, sample_col) {
  ddf_dataframe <- ddf_raw %>% as.data.frame()
  row.names(ddf_dataframe) <- ddf_raw[[sample_col]]
  ddf_dataframe <- ddf_dataframe[sort(rownames(ddf_dataframe)), ]
  ddf_dataframe
}
```

```{r}
ddf$tissue_highfat <- paste(ddf$source, ddf$over_fat_cutoff, sep="_")
ddf_dataframe <- make_deseq_design(ddf, "sampid")
# ddf_dataframe <- make_deseq_design(ddf %>% filter(!is_validation_sample), "sampid")
sdf_dataframe <- make_deseq_data(rdf, rownames(ddf_dataframe))

dds_nocovar <- DESeqDataSetFromMatrix(
  countData = sdf_dataframe,
  colData = ddf_dataframe,
  design = ~tissue_highfat
)

dds_covar <- DESeqDataSetFromMatrix(
  countData = sdf_dataframe,
  colData = ddf_dataframe %>% mutate(smrin=scale(smrin), hardy_scale=make.names(hardy_scale)),
  design = ~tissue_highfat + sex + hardy_scale + smrin
)

dds_out_nocovar <- DESeq(dds_nocovar)
dds_out_covar <- DESeq(dds_covar)

resultsNames(dds_out)

res_liver_nocovar <- results(dds_out_nocovar, contrast=c("tissue_highfat", "liver_TRUE", "liver_FALSE"))
res_pancreas_nocovar <- results(dds_out_nocovar, contrast=c("tissue_highfat", "pancreas_TRUE", "pancreas_FALSE"))

res_liver_covar <- results(dds_out_covar, contrast=c("tissue_highfat", "liver_TRUE", "liver_FALSE"))
res_pancreas_covar <- results(dds_out_covar, contrast=c("tissue_highfat", "pancreas_TRUE", "pancreas_FALSE"))

res_liver_covar %>% data.frame() %>% filter(padj < 0.05 & abs(log2FoldChange) > 1) %>% nrow()
res_pancreas_covar %>% data.frame() %>% filter(padj < 0.05 & abs(log2FoldChange) > 1) %>% nrow()
res_liver_nocovar %>% data.frame() %>% filter(padj < 0.05 & abs(log2FoldChange) > 1) %>% nrow()
res_pancreas_nocovar %>% data.frame() %>% filter(padj < 0.05 & abs(log2FoldChange) > 1) %>% nrow()

# res_liver_shrunk <- lfcShrink(dds_out, contrast=c("tissue_highfat", "liver_TRUE", "liver_FALSE"), res=res_liver)

# res_pancreas_shrunk <- lfcShrink(dds_out, contrast=c("tissue_highfat", "pancreas_TRUE", "pancreas_FALSE"), res=res_pancreas)
```

### P-value histograms

```{r fig.height=6,fig.width=6}
res_liver_nocovar %>% data.frame() %>% ggplot(aes(x=pvalue)) + geom_histogram(bins=100, na.rm = TRUE)
liver_phist <- res_liver_covar %>% data.frame() %>% ggplot(aes(x=pvalue)) + geom_histogram(bins=100, na.rm = TRUE)

res_pancreas_nocovar %>% data.frame() %>% ggplot(aes(x=pvalue)) + geom_histogram(bins=100, na.rm = TRUE)
panc_phist <- res_pancreas_covar %>% data.frame() %>% ggplot(aes(x=pvalue)) + geom_histogram(bins=100, na.rm = TRUE)

head(res_liver_covar)

res_liver_nocovar %>% data.frame() %>% ggplot(aes(x=log2FoldChange, y=-log10(padj))) + geom_point()
res_liver_covar %>% data.frame() %>% ggplot(aes(x=log2FoldChange, y=-log10(padj))) + geom_point()
res_pancreas_nocovar %>% data.frame() %>% ggplot(aes(x=log2FoldChange, y=-log10(padj))) + geom_point()
res_pancreas_covar %>% data.frame() %>% ggplot(aes(x=log2FoldChange, y=-log10(padj))) + geom_point()

panel_plt <- ggarrange(
  res_liver_covar %>% data.frame() %>% ggplot(aes(x=pvalue)) + geom_histogram(bins=100, na.rm = TRUE) + ggtitle("Liver") + xlab("P-value") + ylab("Count") + scale_fill_manual(values=c("#ff8c00")),
  res_pancreas_covar %>% data.frame() %>% ggplot(aes(x=pvalue)) + geom_histogram(bins=100, na.rm = TRUE) + ggtitle("Pancreas") + xlab("P-value") + ylab("Count") + scale_fill_manual(values=c("#ff8c00")),
  res_liver_covar %>% data.frame() %>% ggplot(aes(x=log2FoldChange, y=-log10(padj), color=padj<0.05 & abs(log2FoldChange)>1)) + geom_point(na.rm=TRUE, alpha=0.5) + ylim(0, 12) + xlim(-6, 6) + 
    labs(color="FDR < 0.05, log2 fold > 1") + scale_color_manual(values=c("#888888", "#ff8c00")),
  res_pancreas_covar %>% data.frame() %>% ggplot(aes(x=log2FoldChange, y=-log10(padj), color=padj<0.05 & abs(log2FoldChange)>1)) + geom_point(na.rm=TRUE, alpha=0.5) + ylim(0, 12) + xlim(-6, 6) + 
    scale_color_manual(values=c("#888888", "#ff8c00")), common.legend = TRUE
) %>% ggpubr::annotate_figure(top="Comparison of high / low fat percentage samples")

panel_plt
ggsave(panel_plt, filename = "output/stat_panel_plt.png", width = 8, height = 6)
```





## Writing results prepared for OmicLoupe

```{r}
library(mygene)

all_genes <- getGenes(rdf$Name %>% gsub("\\..", "", .))
all_genes_descr_only <- all_genes %>% data.frame() %>% dplyr::select(query, name)
all_genes_descr_only

adf <- rdf %>% dplyr::select(c("Name", "Description")) %>% 
  dplyr::mutate(query=gsub("\\..", "", Name)) %>%
  inner_join(all_genes_descr_only, by="query") %>%
  dplyr::select(Name, Description, Annot=name) %>%
  distinct(Name, .keep_all = TRUE)
sdf <- rdf %>% dplyr::select(-one_of(c("Name", "Description")))



vsd <- vst(dds_out, blind=FALSE)

out_rdf <- cbind(
  adf,
  res_liver %>% data.frame() %>% rename_all(funs(paste("liver", colnames(res_liver), sep="."))),
  res_liver_shrunk %>% data.frame() %>% rename_all(funs(paste("liver_shrunk", colnames(res_liver), sep="."))),
  res_pancreas %>% data.frame() %>% rename_all(funs(paste("panc", colnames(res_liver), sep="."))),
  res_pancreas_shrunk %>% data.frame() %>% rename_all(funs(paste("panc_shrunk", colnames(res_liver), sep="."))),
  assay(vsd)
)

table(is.na(adf$Annot))
nrow(adf)

dir.create("output")
write_tsv(out_rdf, path = "output/out_rdf.tsv")


set.seed(37)
write_tsv(out_rdf %>% sample_n(1000, replace = FALSE), path = "output/out_rdf_1000.tsv")
write_tsv(out_rdf %>% sample_n(5000, replace = FALSE), path = "output/out_rdf_5000.tsv")
write_tsv(out_rdf, path = "output/out_rdf.tsv")
write_tsv(ddf %>% dplyr::select(sampid, everything()), path = "output/out_rdf.design.tsv")

out_rdf
```

### PCA plots

```{r}
make_pca <- function(pca_obj, fertility, title, text=NULL, pc1=1, pc2=2, padding=1) {
  
  scale_colors <- c("blue", "gray", "red")
  dot_size <- 5
  
  if (is.null(text)) {
    text_col <- "notext"
  }
  else {
    text_col <- text
  }
  
  plt <- ggplot(cbind(pca_obj$rotated, fert_group=fertility, label=text_col), 
                aes_string(x=sprintf("PC%s", pc1), y=sprintf("PC%s", pc2), color="fert_group", label="label")) + 
    xlab(get_load(pca_obj$variance, pc1)) + 
    ylab(get_load(pca_obj$variance, pc2)) + 
    theme_classic() + 
    geom_point(size=dot_size) +
    scale_color_gradientn(colors=scale_colors) +
    ggtitle(title)
  
  if (!is.null(text)) {
    plt <- plt + ggrepel::geom_text_repel(point.padding = padding)
  }
  plt
}

get_pca_matrix <- function(rdf, ddf, sample_name, plot_name) {
  obj <- list()
  obj$matrix <- rdf %>%
    dplyr::select(ddf[[sample_name]]) %>% 
    filter_all(all_vars(!is.na(.))) %>%
    as.matrix()
  obj$pca <- obj$matrix %>% PCAtools::pca(.)
  obj$hc <- hclust(dist(obj$matrix), "ave")
  obj$ddf <- ddf
  obj$plot_name <- plot_name
  obj
}

```

```{r}
liver_ddf <- ddf %>% filter(source == "liver")
panc_ddf <- ddf %>% filter(source == "pancreas")

sdf <- assay(vsd)

liver_sdf <- sdf[, liver_ddf$sampid]
panc_sdf <- sdf[, panc_ddf$sampid]


pca_obj_liver <- PCAtools::pca(liver_sdf, scale=TRUE, center = TRUE, removeVar = 0.5)
pca_obj_panc <- PCAtools::pca(panc_sdf, scale=TRUE, center = TRUE, removeVar = 0.5)

get_load <- function(all, ind) {
  sprintf("PC%s (%s %%)", ind, round(all[ind] / sum(all) * 100, 3))
}

hardy_plt <- ggarrange(
  cbind(pca_obj_liver$rotated %>% data.frame(), liver_ddf) %>% ggplot(aes(x=PC1, y=PC2, color=hardy_scale)) + geom_point(size=2) + xlab(get_load(pca_obj_liver$variance, 1)) + ylab(get_load(pca_obj_liver$variance, 2)) + ggtitle("Liver"),
  cbind(pca_obj_panc$rotated %>% data.frame(), panc_ddf) %>% ggplot(aes(x=PC1, y=PC2, color=hardy_scale)) + geom_point(size=2) + xlab(get_load(pca_obj_panc$variance, 1)) + ylab(get_load(pca_obj_panc$variance, 2)) + ggtitle("Pancreas"),
  common.legend = TRUE, legend="bottom"
)

rin_plt <- ggarrange(
  cbind(pca_obj_liver$rotated %>% data.frame(), liver_ddf) %>% ggplot(aes(x=PC1, y=PC2, color=smrin)) + geom_point(size=2) + xlab(get_load(pca_obj_liver$variance, 1)) + ylab(get_load(pca_obj_liver$variance, 2)) +
    scale_color_gradientn(colors=c("blue", "gray", "red")),
  cbind(pca_obj_panc$rotated %>% data.frame(), panc_ddf) %>% ggplot(aes(x=PC1, y=PC2, color=smrin)) + geom_point(size=2) + xlab(get_load(pca_obj_panc$variance, 1)) + ylab(get_load(pca_obj_panc$variance, 2)) + 
    scale_color_gradientn(colors=c("blue", "gray", "red")),
  common.legend = TRUE, legend="bottom"
)

ggsave(ggarrange(hardy_plt, rin_plt, nrow=2) %>% annotate_figure(top="Principal component analysis, technical factors"), width = 8, height=6, filename = "output/pca_panel.png")
```

```{r}
liver_ddf


liver_ddf$pathology <- liver_ddf$pathology_categories %>% gsub(",.*", "", .)
panc_ddf$pathology <- panc_ddf$pathology_categories %>% gsub(",.*", "", .)

out_plt <- ggarrange(
  cbind(pca_obj_liver$rotated %>% data.frame(), liver_ddf) %>% 
    ggplot(aes(x=PC1, y=PC2, color=pathology)) + geom_point(size=5) + xlab(get_load(pca_obj_liver$variance, 1)) + ylab(get_load(pca_obj_liver$variance, 2)) + ggtitle("Liver") + theme(text=element_text(size=20)) +labs(color="Pathology"),
  cbind(pca_obj_panc$rotated %>% data.frame(), panc_ddf) %>% 
    ggplot(aes(x=PC1, y=PC2, color=pathology)) + geom_point(size=5) + xlab(get_load(pca_obj_panc$variance, 1)) + ylab(get_load(pca_obj_panc$variance, 2)) + ggtitle("Pancreas") + theme(text=element_text(size=20)) +labs(color="Pathology"),
  common.legend = TRUE, legend="bottom", nrow=2
)

ggsave(out_plt, filename = "output/pathology_pca.png", width = 10, height = 10)
```



## Writing shortlists

```{r}
# library(org.Hs.eg.db)
# library(EnsDb.Hsapiens.v79)


liver_shortlist <- out_rdf %>% dplyr::filter(liver.padj < 0.05 & abs(liver.log2FoldChange) > 1) %>% arrange(liver.pvalue)
panc_shortlist <- out_rdf %>% dplyr::filter(panc.padj < 0.05 & abs(panc.log2FoldChange) > 1) %>% arrange(panc.pvalue)

liver_shortlist %>% nrow()
panc_shortlist %>% nrow()

write_tsv(liver_shortlist, path = "output/liver_shortlist.tsv")
write_tsv(panc_shortlist, path = "output/panc_shortlist.tsv")

```


# Explorative visuals

```{r}
res_liver

phist_liv <- res_liver %>% data.frame() %>%
  ggplot(aes(x=pvalue)) + geom_histogram(bins=100)

phist_panc <- res_pancreas %>% data.frame() %>%
  ggplot(aes(x=pvalue)) + geom_histogram(bins=100)

ggsave(phist_liv, filename = "output/phist_liv.png")
ggsave(phist_panc, filename = "output/phist_panc.png")

res_liver_shrunk %>% data.frame() %>%
  ggplot(aes(x=pvalue)) + geom_histogram(bins=100)

res_pancreas_shrunk %>% data.frame() %>%
  ggplot(aes(x=pvalue)) + geom_histogram(bins=100)


```

# Preparing matrices for machine learning

* Select for significant and lowly correlated features using the caret findCorrelation
* Properly parse the two different data frames

```{r}
liver_ids <- ddf %>% dplyr::filter(source == "liver") %>% pull(sampid)
pancreas_ids <- ddf %>% dplyr::filter(source == "pancreas") %>% pull(sampid)


setup_transcript_matrix_for_ml <- function(shortlist_rdf, target_ids, cor_thres=0.5) {
  sdf <- shortlist_rdf %>% dplyr::select(all_of(target_ids))
  corr_matrix <- cor(sdf)
  highly_corr <- findCorrelation(corr_matrix, cutoff=cor_thres)
  low_corr <- sdf[-highly_corr, ]
  low_corr_t <- low_corr %>% 
    t() %>% 
    data.frame() %>% 
    rownames_to_column("rows")
  colnames(low_corr_t) <- c("rows", shortlist_rdf$Description[-highly_corr])
  low_corr_t
}

liver_transcript_ddf <- setup_transcript_matrix_for_ml(liver_shortlist, liver_ids, cor_thres=0.5)
panc_transcript_ddf <- setup_transcript_matrix_for_ml(panc_shortlist, pancreas_ids)

message("All IDs in the transcript ddfs should match the ddf to be bound to, otherwise problems! All matching: ", all(liver_transcript_ddf$rows == ddf %>% dplyr::filter(source == "liver") %>% pull(sampid)))
message("All IDs in the transcript ddfs should match the ddf to be bound to, otherwise problems! All matching: ", all(panc_transcript_ddf$rows == ddf %>% dplyr::filter(source == "pancreas") %>% pull(sampid)))

liver_out <- cbind(ddf %>% dplyr::filter(source == "liver"), liver_transcript_ddf)
pancreas_out <- cbind(ddf %>% dplyr::filter(source == "pancreas"), panc_transcript_ddf)

write_tsv(liver_out, path = "output/liver_design_to_ml.tsv")
write_tsv(pancreas_out, path = "output/pancreas_design_to_ml.tsv")
```










